The state of knowledge for marine invertebrate biodiversity in the continental US

### 2018-01-17 manuscript.Rmd
add sentence about inadequacy of sampling
[fix #84] and other text edits

To assess the level of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] that are available for the United States. For molluscs, of the `r format(itis_stats$moll_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_moll_idig`% are in iDigBio and `r itis_stats$prop_moll_obis`% are in OBIS. For crustaceans, of the `r format(itis_stats$crust_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS. Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (see below), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally. Therefore, we can estimate that there is at least 29,000 species of marine invertebrates in US waters. The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.

@@

Spatial patterns of species richness are however strongly confounded by sampling effort. There is a tight correlation between the number of samples and number of species per cell used in Fig. \@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r format(stat_sampling_effort$n, big.mark=",")`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig. \@ref(fig:correlation-sampling)).

@@

Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data and for the efficient collection of new data.

### Improve existing data

With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text analysis, it is increasingly easier to validate and spell check the scientific names of the identifications of digitized museum specimens. Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability. 
<!-- GP comments: You still have the human knowledge/ID factor to contend with though, which is a more important issue.  Also,places like WoRMS almost never link a given name to multiple useages and useage changes with time -->
Additionally, because registers of species names, such as WoRMS, link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.

@@

For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and provide solutions. There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records (e.g. see [@Otegui2016] for vertebrate records). Such services, however, require that resources be made available to museum collections in order to address these issues directly in collection databases. <!-- GP comment: see geolocate plugin in Specify -->

Increasing the computing literacy and teaching "data wrangling" skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions are incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.

@@

<!-- Finally - need better links in GenBank and BOLD to vouchers, names, and geography -->

<!-- Specimens with accurate identifications and specimens that have been preserved in conditions such that it is possible to get DNA out of them should be prioritized, as they yield the most data that are relevant to biodiversity knowledge. -->

@@

<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it -->

<!-- Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates.  -->

<!-- For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014]. -->

<!-- GP comment: Several of these recommendations match ones made in Duffy et al 2013 -- ref 4  Should emphasize the overlap -->

Museum specimens are critical for validating the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].

This study highlights that we have too little data about too few species of marine invertebrates. We recommend that additional coordinated field expeditions be organized to sample more comprehensively the marine environment. The relationship between sampling effort and diversity from museum collections makes clear that any additional samples will greatly improve our knowledge of marine biodiversity. Concerted and collaborative efforts are needed to share expertise and limited resources.

<!-- GP comment: You could give a biodiversity survey as an example - e.g. in Kaneohe we did ~2400 records of ~1300 species in 11 days - in context of whole US effort, this is substantial, especially at the species level. -->

Samples collected during these expeditions need to be photographed to document live appearance, information that would otherwise be lost during the preservation process. This has been shown to be particularly important to validate species limits in cryptic complexes. We also need to capture metadata associated with specimens to be able to understand better the physical characteristics of the habitats and the life histories of these species. Specimens need to be sub-sampled to create material suitable for genomic analysis. Parts of data acquisition associated with specimens and their biota can be automated and captured digitally directly in the field, speeding up their ingestion in database collections.

Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new discoveries. For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of which taxonomic group would benefit most from additional sampling effort or give an indication of the proportion of species that has not been sampled yet. Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel with metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the methods and permits resolution of specimen identification conflicts.  With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.

@@

DNA sequences are an early test case on how to link biodiversity data across repositories. These initial efforts have revealed the lack of infrastructure for using data formats specifically designed for this task, the lack of ontologies available for biodiversity data, and the fragility of the identifiers [@Stucky2014; @Walls2014; @Guralnick2014]. As new types of data will be collected, added, and linked across repositories, the community needs to develop modern and robust ways to integrate these heterogeneous biodiversity data sources [@Guralnick2015; @Deck2015].

@@

<!-- widening gap with other groups (vertebrates, plants) for which more comprehensive data is available -->

<!-- here there are 4 things: data availability (having good APIs to access the
data); data comprehensiveness (how many species are captured in the database);
data accuracy (can we trust the data that is found in the databases); data linkage -->

@@

>>>>

To assess the level of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] that are available for the United States. For molluscs, of the `r format(itis_stats$moll_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_moll_idig`% are in iDigBio and `r itis_stats$prop_moll_obis`% are in OBIS. For crustaceans, of the `r format(itis_stats$crust_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS. Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (see below), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally. Therefore, using a conservative estimate of a 60% digitization level of the fauna, we can estimate that there is at least 29,000 species of marine invertebrates in US waters. The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.

@@

Spatial patterns of species richness are however strongly confounded by sampling effort. There is a tight correlation between the number of samples and number of species per cell used in Fig. \@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r format(stat_sampling_effort$n, big.mark=",")`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig. \@ref(fig:correlation-sampling)). For most of the US EEZ, the sampling effort is insufficient to capture acurately the species diversity.

@@

Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data and for the efficient collection of new data. Our recommendations extends the recommendations formulated by Duffy et al. [@Duffy2013], and are informed by the results of our analyses.

### Improve existing data quality and access

Duffy et al. [@Duffy2013] suggested that resources should be invested in the digitization of historical biodiversity to study changes in biodiversity through time in response to climate change, fishing pressure, and pollution. Here, we extend this recommendation to also include improving the quality of the data that is already digitized.

With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text analysis, it is increasingly easier to validate and spell check the scientific names of the identifications of digitized museum specimens. Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.

@@

For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and provide solutions. There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records (e.g. see [@Otegui2016] for vertebrate records). Such services, however, require that resources be made available to museum collections in order to address these issues directly in collection databases.

Increasing the computing literacy and teaching "data wrangling" skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions are incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community. As more people know how to analyze data, and in turn, use the data for their research, inaccuracies can be fixed and overall data quality can improve.

@@

@@

This study highlights that we have too little data about too few species of marine invertebrates. We recommend that additional coordinated field expeditions be organized to sample more comprehensively the marine environment. The relationship between sampling effort and diversity from museum collections makes clear that any additional samples will lead to improving our knowledge of species ranges, and will lead to the discovery of many new species. Concerted and collaborative efforts are needed to share expertise and limited resources. Biodiversity surveys that target all animal phyla can be very effective in documenting biodiversity. For instance, one of us (GP) participated in the summer 2017 in a an inventory of Kaneohe Bay, Hawaii, where ~2,400 records representing ~1,300 species were collected in 11 days of field work. While diversity is higher in Hawaii than in the continental US, this is a modest effort that led to the collection of a number of species that represent a few percents of the total diversity of the US EEZ marine invertebrate fauna.

Museum specimens are critical for validating the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000]. Samples collected during biodiversity surveys need to be photographed to document live appearance, information that would otherwise be lost during the preservation process. This has been shown to be particularly important to validate species limits in cryptic complexes. We also need to capture metadata associated with specimens to be able to understand better the physical characteristics of the habitats and the life histories of these species. Specimens need to be sub-sampled to create material suitable for genomic analysis. Parts of data acquisition associated with specimens and their biota can be automated and captured digitally directly in the field, speeding up their ingestion in database collections.

Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new discoveries. For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of the taxonomic groups that would benefit most from additional sampling effort, or provide an estimate for the proportion of species that has not been sampled yet. Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel with metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the methods and permits resolution of specimen identification conflicts.  With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.

@@

To make biodiversity science and conservation decisions more data driven, we need to ensure that biodiversity data is accurate, comprehensive, freely available, and that the datasets are linked. DNA sequences are an early test case on how to link biodiversity data across repositories. These initial efforts have revealed the lack of infrastructure for using data formats specifically designed for this task, the lack of ontologies available for biodiversity data, and the fragility of the identifiers [@Stucky2014; @Walls2014; @Guralnick2014]. As new types of data will be collected, added, and linked across repositories, the community needs to develop modern and robust ways to integrate these heterogeneous biodiversity data sources [@Guralnick2015; @Deck2015].

@@

@@

Species occurence data and museum records were extracted from OBIS (Ocean Biogeographic Information System) and iDigBio (Integrated Digitized Biocollections) respectively. OBIS gathers and integrates datasets about occurrences of marine species around the world provided by over 500 institutions worldwide. iDigBio provides access to the data associated with museum records that have been digitized and made publicly available. While OBIS records can be based on observation, there is a physical specimen deposited in a collection associated with each iDigBio record.
